Apply BioQC on samples with tissues annotated
=============================================

This is our first bulk analysis running BioQC on all GEO datasets (GDS) 
that have a tissue annotated. 

```{r load_libries}
library(pylr)
library(parallel)
library(tools)
source("lib.R")
eset.path = "/homebasel/biocomp/sturmg/projects/GEO_BioQC/data/GDS_all/%s.Rdata"
bioqc.res.path = "/homebasel/biocomp/sturmg/projects/GEO_BioQC/data/GDS_all/plots/%s_bioqc_res.tab"
studies = sapply(Sys.glob(sprintf(eset.path, "*")), function(x) {return (file_path_sans_ext(basename(x)))})
```

1. We download all GDS from GEO using chunksub
2. We have a look at the available attributes in the pData(eset)

```{r}
getAttributes = function(study) { 
  load(sprintf(eset.path, study))
  return(list(colnames(pData(eset))))
}
r = unlist(mcmapply(getAttributes, studies, mc.cores=40))
c = count(r)
attributes = c[with(c, order(-freq)),]
write.table(attributes, file="/homebasel/biocomp/sturmg/projects/GEO_BioQC/BioQC_GEO_analysis/results/geo_gds_attributes.tsv")
```

3. We will now get an overview over the abundance of the respective tissues, so that we can choose the top $n$ tissues on which we can then perform the contamination analysis. 

```{r}
getTissues = function(study) {
  load(sprintf(eset.path, study))
  if("tissue" %in% colnames(pData(eset))) {
    return(list(lapply(lapply(pData(eset)$tissue, as.character), tolower)))
  } else {
    print(paste(basename(study), "does not contain any tissue annotation. "))
    return(list(c("no_tissue_annotation")))
  }
}
r = unlist(mcmapply(getTissues, studies, mc.cores=40))
c = count(r)
tissues = c[with(c, order(-freq)),]
write.table(tissues, file="/homebasel/biocomp/sturmg/projects/GEO_BioQC/BioQC_GEO_analysis/results/geo_gds_tissues.tsv")
```

4. We now merge all samples from one tissue into one expression set.
We can then run BioQC on this Expression set and identify contamined examples. 

```{r}
# init samplesPerTissue list. 
#
# We create a list (used as dictionary) to store a dataframe 
# for each tissue that will contain the respective BioQC results. 
tissueList = as.character(tissues[,1])
gmtFile = system.file("extdata/exp.tissuemark.affy.roche.symbols.gmt", package="BioQC")
gmt <- readGmt(gmtFile)
emptySignatureDf = data.frame(row.names=names(gmt))
samplesPerTissue = rep(list(edf), length(tissueList))
names(samplesPerTissue) = tissueList

# extract tissue specific results
#
# Go through all studies in parallel. Return a list 
# tissue -> DataFrame(resultsForTissue) for each 
# study. 
bioQCTissue = function(study) {
  load(sprintf(eset.path, study))
  bioqcRes = read.table(sprintf(bioqc.res.path, study))
  if ("tissue" %in% colnames(pData(eset))) {
    study.tissues = unique(sapply(sapply(pData(eset)$tissue, as.character), tolower))
    getTissueBioqcRes = function(tissue) {
      cols = which(pData(eset)$tissue == tissue)
      res = bioqcRes[, cols, drop=FALSE]
      colnames(res) = sapply(colnames(res), function(x) {return(paste(study, x, sep="."))})
      return(res)
    }
    return(sapply(study.tissues, getTissueBioqcRes))
  } 
}
r.per.study = mcmapply(bioQCTissue, studies, mc.cores=40)

# reduce the per-study results into the samplesPerTissue list. 
for(study in r.per.study) {
  for(tissue in names(study)) {
    samplesPerTissue[[tissue]] = cbind(samplesPerTissue[[tissue]], study[[tissue]])
  }  
}
```


Make a heatmap for each tissue displaying only significant signatures (BioQC score >6 per row)
```{r}
for (tissue in tissueList) {
  print(tissue)
  bioqcResFil <- filterPmat(samplesPerTissue[[tissue]], 1E-6)
  if(ncol(bioqcResFil) > 20 && nrow(bioqcResFil) > 1) {  
    bioqcAbsLogRes <- as.matrix(absLog10p(bioqcResFil))
    bioqcAbsLogResNorm = apply(bioqcAbsLogRes, 2, norm01)
    pdf(file=sprintf("results/heatmaps/%s.pdf", make.names(tissue, unique=TRUE)), width=100, height=15)
    print(bioqcHeatmap(bioqcAbsLogResNorm, title=tissue))
    dev.off()
  }  
}

```

Make a heatmap for each tissue using only the siginificant samples. (BioQC score >6 per column and per column)
```{r}
for (tissue in tissueList) {
  print(tissue)
  bioqcResFil = t(filterPmat(t(samplesPerTissue[[tissue]]), 1E-6))
  bioqcResFil <- filterPmat(bioqcResFil, 1E-6)
  if(ncol(bioqcResFil) > 15 && nrow(bioqcResFil) > 1) {  
    bioqcAbsLogRes <- as.matrix(absLog10p(bioqcResFil))
    bioqcAbsLogResNorm = apply(bioqcAbsLogRes, 2, norm01)
    pdf(file=sprintf("results/heatmaps_sig/%s.pdf", make.names(tissue, unique=TRUE)), width=100, height=15)
    print(bioqcHeatmap(bioqcAbsLogResNorm, title=tissue))
    dev.off()
  }  
}

```

Which are the (second/third/...) highest scoring signatures for each tissue?
```{r}
my.tissues = c("bone marrow", "liver", "kidney", "lung", "colon", "jejunum")

plotTop5 = function(sig.counts, i) {
  # get the top 5 tissues that score #1
  top5 = data.frame(sig.counts[order(sig.counts[,i], decreasing=TRUE)[1:5], i, drop=FALSE])
  # prepare for plotting
  top5['signature'] = rownames(top5)
  top5.molten = melt(top5, id.vars="signature", value.name="count", variable.name="rank" )
  # plot bar chart 
  return(ggplot(top5.molten, aes(x=rank, y=count, fill=signature)) +
          geom_bar(aes(fill=signature), stat="identity", position="dodge") +
          ggtitle(paste(tissue, "#", i)))
}

for (tissue in my.tissues) {
  tissue.scores = samplesPerTissue[[tissue]]
  # filter for significant samples (filter for cols, not rows in that case)
  tissue.scores.fil = t(filterPmat(t(tissue.scores), 1E-6))
  tissue.scores.fil.log = absLog10p(tissue.scores.fil)
  
  sig.ranks = apply(tissue.scores.fil.log, 2, function(x) {return(rank(x, ties.method="first"))})
  sig.counts = data.frame(sapply(1:10, function(i) {
      apply(sig.ranks, 1, function(x){sum(x == nrow(sig.ranks)-i+1)})
    }))
  for (i in 1:5) {
    print(plotTop5(sig.counts, i))
  }
  
}

```

Which tissue signatures are enriched in a certain tissue, even though the signatures show no/little overlap? 
This could show that certain cell types are abundant in multiple tissues. 
-> make a boxplot for the top5 enriched signatures (by median)
```{r}
tissue = "kidney"
tissue.scores = samplesPerTissue[[tissue]]
tissue.scores.log = absLog10p(tissue.scores)
top5.ids = order(apply(tissue.scores.log, 1, median), decreasing=TRUE)[1:5]
top5 = tissue.scores.log[top5.ids, ]
top5['signature'] = rownames(top5)
top5.molten = melt(top5, id.vars="signature", value.name="score", variable.name="study")
ylim1 = boxplot.stats(top5.molten$score)$stats[c(1,5)]
ggplot(top5.molten, aes(x=signature, y=score)) + geom_boxplot() + coord_cartesian(ylim = ylim1*2)

```






