Automatically detect contaimined samples
========================================

Sandbox for testing different methods for identifying  
contamination automatically. 


```{r}


studies = Sys.glob("/homebasel/biocomp/sturmg/projects/GEO_BioQC/GDS_tissue/plots/*_bioqc_res.tab")
contamResDf = as.data.frame(matrix(vector(), nrow=length(studies), ncol=3))
colnames(contamResDf) = c("study", "contamined", "samples")
findContamStudies = function(study) {
    bioqcRes = read.table(study)
    bioqcResFil <- filterPmat(bioqcRes, 1E-6)
    bioqcAbsLogRes <- as.matrix(absLog10p(bioqcResFil))
    contamined = FALSE
    contam.sample.names = ""
    if(dim(bioqcAbsLogRes)[1] > 0) {
       contam.samples = detectContamZ(bioqcAbsLogRes)
       contam.samples = which(apply(contam.samples, 1, sum) > 0)
       contam.sample.names = paste(names(contam.samples), sep=",")
       contamined = length(contam.samples) > 0
    }   
    return(list(contamined=contamined, samples=contam.sample.names))
}
contamStudies = sapply(studies, findContamStudies)
write.table(t(contamStudies), sep='\t', file="/homebasel/biocomp/sturmg/projects/GEO_BioQC/GDS_tissue/contamined.z.tsv")
```


Z-test based
(more than n times the average/median)
```{r}
# Detect Contaminated Samples based on z-score. 
# Returns:
#     boolean matrix sample/tissue with TRUE for contamined. 
detectContamZ = function(bioqcAbsLogRes) {
  thres.z = 6
  contam.samples = apply(bioqcAbsLogRes, 1, function(row) {
    x = mean(row)
    return(row > 6*x)
  })
  return(contam.samples)
}


```


Double Z-test based
(more than n times the average/median in one tissue and 
less than n times the average/median in another tissue)
```{r}
```

Enrichment based. Define a tissue as enriched. If a sample is enriched in a non-enriched tissue, 
define it as contamined. 
```{r}
thres.enriched = 6 # log 10 pval
thres.contam = 15
enriched.tissues = apply(bioqcAbsLogRes, 1, function(row) {
  return (median(row) > thres.enriched) 
})
contam.samples = apply(bioqcAbsLogRes, 1, function(row) {
  enriched = median(row) > thres.enriched
  return(!enriched & (row > thres.contam))
})
```

T test based. (actually we need sth different.) Make a boolean matrix: tissue is overrepresented in sample in comparison to other samples. Make second boolean matrix: tissue is underrepresented in sample in comparison to other samples. If a sample has both an over and an underrepresented tissue, it is likely to be contamined. 
```{r}
test.greater = matrix(False, nrow=nrow(bioqcAbsLogRes), ncol=(bioqcAbsLogRes))
test.less = matrix(False, nrow=nrow(bioqcAbsLogRes), ncol=(bioqcAbsLogRes))
for i in 1:nrow(test.greater) {
  for j in 1:ncol(test.greater) {
    
  }
}
```