# Sample Selection and processing {#sample-selection}

```{r setup, include=FALSE}
source("lib/knitr.R")
library(BioQC)
library(rworldmap)
library(VennDiagram)
library(data.table)
library(grid)
library(gridBase)
library(stringr)
library(testthat)
source("lib/plots.R")
source("lib/db.R")
```

## Sample Preselection
Here, we document the sample selection process before running *BioQC*. 

### Required annotation
A sample is *usable* for this study, if 

* the gene symbos are annotated (requirement to run BioQC)
* the tissue of origin is annotated (requirment to draw conclusions about contamination)

We consider two approaches for annotating gene symbols: 

* Using the Bioconductor [AnnotationDbi](https://bioconductor.org/packages/release/bioc/html/AnnotationDbi.html) package. The GEOmetadb provides a mapping of the GPL identifier to these packages. 
* Using the `annotGPL=TRUE` option of [GEOquery](https://bioconductor.org/packages/release/bioc/html/GEOquery.html)'s `getGEO`. This requires an annotation file being available for the respective platform. We retrieved a list of the available annotation files [in an earlier step](#load-annotation-information). 

We compare the two methods with respect to the amount of usable samples that we can get. 

We create the 'annotation statistics' using an  [sql script](https://github.com/grst/BioQC_GEO_analysis/blob/master/db/views/annotation_stats.sql) and calculate these Venn diagrams: 

 
```{r sample_filtering, echo=FALSE, fig.height=3, fig.width=8, cache=TRUE}
sqlStats = "
select 'total', count(distinct gsm), count(distinct gse) 
from bioqc_studies_total
union
  select 'tissue', count(distinct gsm), count(distinct gse) 
  from bioqc_studies_has_tissue
union 
  select 'package', count(distinct gsm), count(distinct gse)
  from bioqc_studies_has_package
union
  select 'annotation', count(distinct gsm), count(distinct gse)
  from bioqc_studies_has_annot
union 
  select 'tissue_package', count(distinct gsm), count(distinct gse)
  from (
    select *
    from bioqc_studies_has_tissue
    intersect
      select *
      from bioqc_studies_has_package) u1
union
  select 'annotation_package', count(distinct gsm), count(distinct gse) 
  from (
    select *
    from bioqc_studies_has_annot
    intersect
      select *
      from bioqc_studies_has_package) u2
union
  select 'tissue_annotation', count(distinct gsm), count(distinct gse)
  from (
    select *
    from bioqc_studies_has_tissue
    intersect
      select * 
      from bioqc_studies_has_annot) u3
union
  select 'annotation_tissue_package', count(distinct gsm), count(distinct gse)
  from (
    select *
    from bioqc_studies_has_tissue 
    intersect
      select *
      from bioqc_studies_has_package
    intersect
      select *
      from bioqc_studies_has_annot) u4
"
stats = data.table(dbGetQuery(mydb, sqlStats))
colnames(stats) = c("rn", "GSM", "GSE")
setkey(stats,rn)

categories = c("tissue", "annot pkg", "annot GPL")
colors=c('red', 'green', 'blue')

plot.new()
gl = grid.layout(nrow=1, ncol=2)
vp.1 = viewport(layout.pos.col=1, layout.pos.row=1)
vp.2 = viewport(layout.pos.col=2, layout.pos.row=1)
pushViewport(viewport(layout=gl))
pushViewport(vp.1)

invisible(draw.triple.venn(stats[rn=='tissue',GSM], 
                 stats[rn=='package',GSM], 
                 stats[rn=='annotation',GSM], 
                 stats[rn=='tissue_package',GSM],
                 stats[rn=='annotation_package',GSM], 
                 stats[rn=='tissue_annotation',GSM], 
                 stats[rn=='annotation_tissue_package',GSM], 
                 category=categories, fill=colors))

popViewport()
pushViewport(vp.2)

invisible(draw.triple.venn(stats[rn=='tissue',GSE], 
                 stats[rn=='package',GSE], 
                 stats[rn=='annotation',GSE], 
                 stats[rn=='tissue_package',GSE],
                 stats[rn=='annotation_package',GSE], 
                 stats[rn=='tissue_annotation',GSE], 
                 stats[rn=='annotation_tissue_package',GSE],
                 category=categories, fill=colors))

popViewport(1)
```

The `getGEO` method appears to be the more powerful method. Ideal would be a combination of the two, however, for the sake of simplicity, we stick to `getGEO`, loosing `r as.integer(stats[rn=="tissue_package",GSE]-stats[rn=="annotation_tissue_package",GSE])` studies (`r as.integer(stats[rn=="tissue_package",GSM]-stats[rn=="annotation_tissue_package",GSM])` samples). 

This leaves us with the following filtering result: 
```{r sample_filtering2, cache=TRUE, echo=FALSE}
tab = stats[.(c("total", "tissue", "annotation", "tissue_annotation"))]
tab = tab[,comment:=c("total", "tissue annotated", "annotation file available", "tissue and annotation file")]
kable(tab[,c("comment","GSM","GSE"),with=FALSE])
```

```{r gse_path, include=FALSE}
gse.file = "results/gse_lists/gse_tissue_annotation.txt"
```

### Export list of samples {#sample-list}
We store the respective gse identifiers in ``r gse.file``: 
```{r sample_filtering_out}
sqlGse = "
select distinct gse
from (
  select *
  from bioqc_studies_has_tissue
  intersect
    select *
    from bioqc_studies_has_annot) u"
gse = dbGetQuery(mydb, sqlGse)
writeLines(gse$GSE, file(gse.file))
```

