```{r}
library(stringr)
library(readr)
library(MASS)
source("config.R")
library(cowplot)
library(ggplot)
library(boot)
library(dplyr)
select = dplyr::select
```


# Figures for publication

```{r load_data, include=FALSE}
load('../results/data_processed.RData')
load('../results/models.RData')

data_corr_array = data_corr
total_array = data_corr$GSM %>% unique() %>% length()

load('../results/archs4/archs4_data_processed.RData')
load('../results/archs4/archs4_models.RData')

data_corr_ngs = data_corr
total_ngs = data_corr$GSM %>% unique() %>% length()

```

Same as in the section before, but all bioqc signatures aggregated by tissue groups.
Additional, we define samples as being "severely heterogeneous", if the reference signature,
i.e. the signature that should be present according to the annotation, is not enriched at
an unadjusted p-value < `r SEVERE_THRES`.
```{r, include=FALSE}
bioqc_tissue_set = read_csv("../data/bioqc_geo_oracle_dump/BIOQC_TISSUE_SET_DATA_TABLE.csv") %>%
  filter(TISSUE_SET == 'bioqc_all') %>%
  select(SIG_ID=SIGNATURE, DETECTED_TGROUP=TGROUP)

bioqc_all = selected_signatures %>%
  filter(str_detect(SIGNATURE, 'BIOQC')) %>%
  pull("SIGNATURE")

contam_ngs_tgroup = data_corr_ngs %>%
  inner_join(bioqc_signatures, by=c("SIGNATURE"="SIG_NAME")) %>%
  rename(SIG_ID=ID) %>%
  inner_join(bioqc_tissue_set) %>%
  mutate(is_contaminated=qvalue < TAU & q_corr < TAU & TGROUP != DETECTED_TGROUP) %>%
  mutate(is_severe=is_contaminated & REF_PVALUE >= SEVERE_THRES) %>%
  mutate(dataset = "ARCHS4")

contam_array_tgroup = data_corr_array %>%
  inner_join(bioqc_signatures, by=c("SIGNATURE"="SIG_NAME")) %>%
  rename(SIG_ID=ID) %>%
  inner_join(bioqc_tissue_set) %>%
  mutate(is_contaminated=qvalue < TAU & q_corr < TAU & TGROUP != DETECTED_TGROUP) %>%
  mutate(is_severe=is_contaminated & REF_PVALUE >= SEVERE_THRES) %>%
  mutate(dataset = "GEO")

contam_all = bind_rows(contam_ngs_tgroup, contam_array_tgroup) %>% ungroup()
```

```{r export_files, include=FALSE}
heterogeneity_res = contam_all %>%
  select(dataset, GSM, TGROUP, REF_SIG, DETECTED_TGROUP, SIGNATURE, REF_PVALUE, PVALUE, qvalue, p_corr, q_corr, is_contaminated, is_severe) %>%
  distinct()

heterogeneity_res %>%
  write_tsv("../results/heterogeneity_results.tsv")

heterogeneity_res %>%
  select(dataset, GSM, TGROUP, DETECTED_TGROUP, is_contaminated, is_severe) %>%
  distinct() %>%
  write_tsv("../results/heterogeneity_results_summary.tsv")
```

```{r, fig.width=8, fig.height=5, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Absolute number of samples per tissue in which certain signature is detected as significantly enriched. "}
tgroups = bioqc_tissue_set$DETECTED_TGROUP %>% unique()
p2 = contam_all %>%
  filter(is_contaminated == TRUE) %>%
  group_by(dataset, TGROUP, DETECTED_TGROUP) %>%
  summarise(cnt=n_distinct(GSM)) %>%
  ggplot(aes(x=factor(DETECTED_TGROUP, levels=tgroups), y=TGROUP)) +
      geom_tile(aes(fill=cnt)) +
      geom_text(aes(label=cnt), size=3) +
      theme_bw() +
      theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=.5, size=10)) +
      scale_x_discrete(drop=FALSE) +
      ylab("reference tissue") +
      xlab("detected signature") +
      background_grid() +
      facet_grid(dataset~.) +
      scale_fill_distiller(palette = "Blues", direction=1, guide=FALSE)

```


We use bootstrapping (R package `boot`) to derive confidence intervals.
```{r, include=FALSE}
stat_fun_contam = function(df, i) {
  df = df[i, ]
  total = nrow(df)
  contaminated = sum(df$sample_is_contaminated)
  contaminated/total
}

stat_fun_severe = function(df, i) {
  df = df[i, ]
  total = nrow(df)
  contaminated = sum(df$sample_is_severe)
  contaminated/total
}

bootstrap = function(df_group) {
  boot_obj_contam = boot(df_group, statistic=stat_fun_contam, R=999)
  ci_comtam = boot.ci(boot_obj_contam, conf=0.95, type="basic")
  boot_obj_severe = boot(df_group, statistic=stat_fun_severe, R=999)
  ci_severe = boot.ci(boot_obj_severe, conf=0.95, type="basic")
  tibble(type=c("heterogeneity", "severe heterogeneity"),
         fraction=c(boot_obj_contam$t0, boot_obj_severe$t0),
         ci_low = c(ci_comtam$basic[1,4], ci_severe$basic[1,4]),
         ci_high = c(ci_comtam$basic[1, 5], ci_severe$basic[1,5]))
}

contam_frac = contam_all %>%
  bind_rows(contam_all %>% mutate(TGROUP = "all")) %>%
  group_by(dataset, TGROUP, GSM) %>%
  summarise(sample_is_contaminated=any(is_contaminated), sample_is_severe=any(is_severe)) %>%
  group_by(dataset, TGROUP) %>%
  do(bootstrap(.))
```

```{r, fig.width=8, fig.height=4.5, echo=FALSE, warning=FALSE, fig.cap="fraction of heterogenous samples per tissue. "}
p1 = contam_frac %>%
  ungroup() %>%
  ggplot(aes(y=fraction, x=TGROUP)) +
    geom_bar(aes(fill=dataset), stat="identity", position = "dodge") +
    geom_errorbar(aes(ymin=if_else(ci_low < 0, 0, ci_low), ymax=ci_high, fill=dataset), position='dodge') +
    scale_fill_brewer(palette="Set1") +
    theme(axis.text.x = element_text(angle = 90, hjust=1, vjust=.5),
          legend.position = "top",
          axis.title=element_blank()) +
    facet_wrap(~type) +
    ylab("frac. heterogeneous samples")

ggsave("../results/figures/heterogeneity_fractions.pdf")
ggsave("../results/figures/heterogeneity_fractions.png")
write_tsv(contam_frac, path = "../results/heterogeneity_fractions.tsv")
```

```{r, fig.width=8, fig.height=9}
plot_grid(p1, p2, align = "v", nrow=2, rel_heights = c(.5, .6), labels = "AUTO", axis="lr")
ggsave("../results/figures/heterogeneity_main.pdf")
ggsave("../results/figures/heterogeneity_main.png")
```
